---
title: "Project, STAT 633A"
author: "Wasima Alam"
format: 
   html:
      self-contained: true
echo: true
---

## 

```{r}
library(tidyverse)
library(dplyr)
library(janitor)
library(car)
library(faraway)
```

```{r}
life_exp <- read.csv("Life Expectancy Data (1).csv")
head(life_exp)

# clean column names 
life_exp <- clean_names(life_exp)

life_exp2 <- life_exp |>
  filter(year == 2015)
summary(life_exp2)

# check for missing values - NAs in Alcohol (177), Hepatitis B (9), BMI (2), Total expenditure (181), GDP (29), Population (41), Thinness 1 19 (2), Thinness 5 9 (2), Income (10), Schooling (10)

life_exp3 <- life_exp2 |>
  select(-alcohol, -total_expenditure, -percentage_expenditure)


life_exp4 <- na.omit(life_exp3)
```

```{r}

# Make status a dummy variable
life_exp4$status <- as.factor(life_exp4$status)

# Check reference level
levels(life_exp4$status)

# Fit multiple regression model

mlr_model <- lm(
  life_expectancy ~ status + adult_mortality + infant_deaths + hepatitis_b + measles + bmi +
    under_five_deaths + polio + diphtheria + hiv_aids + population +
    gdp + thinness_1_19_years + thinness_5_9_years +
    income_composition_of_resources + schooling,
  data = life_exp4
)

summary(mlr_model)
```

**Strong significant predictors (p \< 0.05)**

**Marginally significant (0.05--0.10)**

**Not significant (p \> 0.05)**

Because the p-value is much smaller than 0.05, we reject the null hypothesis. There is strong evidence that life expectancy is related to at least one predictor in your model.

```{r}
# Scatter plot
pairs(
   life_expectancy ~ status + adult_mortality + infant_deaths + hepatitis_b + measles + bmi + population +
    under_five_deaths + polio + diphtheria + hiv_aids +
    gdp + thinness_1_19_years + thinness_5_9_years +
    income_composition_of_resources + schooling,
  data = life_exp4
)


```

```{r}
# Boxplot for dummy variable

boxplot(life_expectancy ~ status,
        data = life_exp4,
        main = "Life Expectancy by Status",
        xlab = "Status",
        ylab = "Life Expectancy")
```

Developed countries live longer and more consistently than developing countries.

**Potential problems:**

```{r}
# standardized residuals
r <- rstandard(mlr_model)

# fitted values
yhat <- fitted(mlr_model)

# residual plot
plot(yhat, r,
     xlab = "Fitted values",
     ylab = "Standardized residuals",
     main = "Standardized Residuals vs Fitted")
abline(h = 0, lty = 2)

# QQ plot
qqnorm(r)
qqline(r)

# Shapiro-Wilk
shapiro.test(resid(mlr_model))

# outliers
ind <- which(abs(r) > 2)
life_exp4[ind, ]



# leverage
h <- hatvalues(mlr_model)
n <- nrow(life_exp4)
p <- length(coef(mlr_model)) - 1
cutoff_leverage <- 2 * (p + 1) / n
which(h > cutoff_leverage)

# VIF
library(car)

vif(mlr_model)
```

```{r}
pacman::p_load(MASS, car)
summary(powerTransform(mlr_model))
boxcox(mlr_model, lambda=seq(0.3, 0.65, by=0.05))
```

After doing the predictor variables transformation then will do the boxcox for Y transformation.

```{r}
step_model <- step(mlr_model, trace = 0)

summary(step_model)
AIC(step_model)
```

### Scatter plots - Response vs predictors

```{r}
plot(life_expectancy ~ adult_mortality, data = life_exp4)
```

```{r}
plot(life_expectancy ~ hepatitis_b, data = life_exp4)

```

```{r}
plot(life_expectancy ~ measles, data = life_exp4)
plot(life_expectancy ~ log(measles + 1), data = life_exp4)
```

```{r}
plot(life_expectancy ~ bmi, data = life_exp4)
```

```{r}
plot(life_expectancy ~ under_five_deaths, data = life_exp4)
plot(life_expectancy ~ log(under_five_deaths + 1), data = life_exp4)
```

```{r}
plot(life_expectancy ~ polio, data = life_exp4)
```

```{r}
plot(life_expectancy ~ hiv_aids, data = life_exp4)
plot(life_expectancy ~ log(hiv_aids), data = life_exp4)
```

```{r}
plot(life_expectancy ~ population, data = life_exp4)
plot(life_expectancy ~ log(population), data = life_exp4)
```

```{r}
plot(life_expectancy ~ thinness_5_9_years, data = life_exp4)

```

```{r}
plot(life_expectancy ~ schooling, data = life_exp4)
```

```{r}
plot(life_expectancy ~ infant_deaths, data = life_exp4)
plot(life_expectancy ~ log(infant_deaths + 1), data = life_exp4)
```

```{r}
plot(life_expectancy ~ diphtheria, data = life_exp4)

```

```{r}
plot(life_expectancy ~ gdp, data = life_exp4)
plot(life_expectancy ~ log(gdp), data = life_exp4)
```

```{r}
plot(life_expectancy ~ thinness_1_19_years, data = life_exp4)
```

```{r}
plot(life_expectancy ~ income_composition_of_resources, data = life_exp4)
```

### Standardized Residuals vs predictors

```{r}
# constant variance satisfied
plot(life_exp4$adult_mortality, rstandard(mlr_model),xlab = "Adult Mortality", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plotlog(life_exp4$infant_deaths, rstandard(mlr_model),xlab = "infant death", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$adult_mortality, rstandard(mlr_model),xlab = "Adult Mortality", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$percentage_expenditure, rstandard(mlr_model),xlab = "percentage expenditure", ylab = "Standardized Residuals")
plot(log(life_exp4$percentage_expenditure), rstandard(mlr_model),xlab = "percentage expenditure", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$hepatitis_b, rstandard(mlr_model),xlab = "hepatitis_b", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$measles, rstandard(mlr_model),xlab = "measles", ylab = "Standardized Residuals")
plot(log(life_exp4$measles), rstandard(mlr_model),xlab = "Measles", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$bmi, rstandard(mlr_model),xlab = "bmi", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$under_five_deaths, rstandard(mlr_model),xlab = "under_five_deaths", ylab = "Standardized Residuals")
plot(log(life_exp4$under_five_deaths), rstandard(mlr_model),xlab = "under five deaths", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$polio, rstandard(mlr_model),xlab = "polio", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$diphtheria, rstandard(mlr_model),xlab = "diphtheria", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$hiv_aids, rstandard(mlr_model),xlab = "hiv_aids", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$gdp, rstandard(mlr_model),xlab = "gdp", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied
plot(life_exp4$thinness_1_19_years, rstandard(mlr_model),xlab = "thinness_1_19", ylab = "Standardized Residuals")

```

```{r}
lm_full <- lm(life_expectancy ~ status + adult_mortality + hepatitis_b + log(measles + 1) + bmi + log(under_five_deaths + 1) + polio + log(hiv_aids) + log(population) + thinness_5_9_years + schooling + log(infant_deaths + 1) + diphtheria + log(gdp) + thinness_1_19_years + income_composition_of_resources + adult_mortality:status + hepatitis_b:status + log(measles + 1):status + bmi:status + log(under_five_deaths + 1):status + polio:status + log(hiv_aids):status + log(population):status + thinness_5_9_years:status + schooling:status + log(infant_deaths + 1):status + diphtheria:status + log(gdp):status + thinness_1_19_years:status + income_composition_of_resources:status, data = life_exp4)

summary(lm_full)
```

```{r}
pacman::p_load(MASS, car)
summary(powerTransform(lm_full))
boxcox(lm_full, lambda=seq(0.7, 3, by=0.05))
```

```{r}
model_step <- step(lm_full, trace = 0)

summary(model_step)
```

```{r}
lm_final <- lm(life_expectancy ~ status + adult_mortality + hepatitis_b + log(measles + 1) + log(hiv_aids) + status:adult_mortality, data = life_exp4)
summary (lm_final)

```

```{r}
# standardized residuals
r <- rstandard(lm_final)

# fitted values
yhat <- fitted(lm_final)

# residual plot
plot(yhat, r,
     xlab = "Fitted values",
     ylab = "Standardized residuals",
     main = "Standardized Residuals vs Fitted")
abline(h = 0, lty = 2)

# QQ plot
qqnorm(r)
qqline(r)

# Shapiro-Wilk
shapiro.test(resid(lm_final))

# outliers
ind <- which(abs(r) > 2)
life_exp4[ind, ]



# leverage
h <- hatvalues(lm_final)
n <- nrow(life_exp4)
p <- length(coef(lm_final)) - 1
cutoff_leverage <- 2 * (p + 1) / n
which(h > cutoff_leverage)

# VIF
library(car)

vif(lm_final)
```

```{r}

```
