---
title: "Project, STAT 633A"
author: "Wasima Alam, Jennifer Soper, Gizem Ozyildirim"
format: 
   html:
      self-contained: true
echo: true
message: false
warning: false
---

## 

```{r}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(janitor)
library(car)
library(faraway)
library(MASS)
library(MuMIn)
```

### Data Cleaning

```{r}

# Data Cleaning Process
life_exp <- read.csv("life_expectancy.csv")
head(life_exp)

# clean column names 
life_exp <- clean_names(life_exp)

# year 2015 only
life_exp2 <- life_exp |>
  filter(year == 2015)

# check for missing values - NAs in Alcohol (177), Hepatitis B (9), BMI (2), Total expenditure (181), GDP (29), Population (41), Thinness 1 19 (2), Thinness 5 9 (2), Income (10), Schooling (10)
summary(life_exp2)

# remove variables due to high number of missing values
# percent expenditure is a function of total expenditure with many zeros
life_exp3 <- life_exp2 |>
  dplyr::select(-alcohol, -total_expenditure, -percentage_expenditure)

# remove remaining rows with missing values - 130 observations total
life_exp4 <- na.omit(life_exp3)
```

### Full Additive Multiple Regression Model

```{r}

# Make status a dummy variable
life_exp4$status <- as.factor(life_exp4$status)

# Check reference level
levels(life_exp4$status)

# Developed = 0, developing = 1
contrasts(life_exp4$status)

# Fit full additive multiple regression model
mlr_model <- lm(
  life_expectancy ~ status + adult_mortality + infant_deaths + hepatitis_b + measles + bmi +
    under_five_deaths + polio + diphtheria + hiv_aids + population +
    gdp + thinness_1_19_years + thinness_5_9_years +
    income_composition_of_resources + schooling,
  data = life_exp4
)

summary(mlr_model)
```

### VIF - potential multicollinearity issues

```{r}
VIF <- round(vif(mlr_model), 2)
vif <- as.data.frame(VIF)
vif

# VIF > 5 
which(round(vif(mlr_model), 2) > 5)

# VIF < 5
which(round(vif(mlr_model), 2) < 5)
```

### Exploratory Visualization of the Relationship Between Life Expectancy and Other Predictors

```{r}
# Scatter plot - not used because too difficult to see relationships
pairs(
   life_expectancy ~ status + adult_mortality + infant_deaths + hepatitis_b + measles + bmi + population +
    under_five_deaths + polio + diphtheria + hiv_aids +
    gdp + thinness_1_19_years + thinness_5_9_years +
    income_composition_of_resources + schooling,
  data = life_exp4
)


```

```{r}
# Boxplot for dummy variable 
boxplot(life_expectancy ~ status,
        data = life_exp4,
        main = "Life Expectancy by Status",
        xlab = "Status",
        ylab = "Life Expectancy")
```

Median life expectancy higher with less variability for developed countries vs developing countries.

### Individual Scatter plots of response vs predictors - exploring the use of transformations on predictors

```{r}
plot(life_expectancy ~ adult_mortality, data = life_exp4)
```

```{r}
plot(life_expectancy ~ hepatitis_b, data = life_exp4)

```

```{r}
# plot(life_expectancy ~ measles, data = life_exp4)
plot(life_expectancy ~ log(measles + 1), data = life_exp4)
```

```{r}
plot(life_expectancy ~ bmi, data = life_exp4)
```

```{r}
# plot(life_expectancy ~ under_five_deaths, data = life_exp4)
plot(life_expectancy ~ log(under_five_deaths + 1), data = life_exp4)
```

```{r}
plot(life_expectancy ~ polio, data = life_exp4)
```

```{r}
# plot(life_expectancy ~ hiv_aids, data = life_exp4)
plot(life_expectancy ~ log(hiv_aids), data = life_exp4)
```

```{r}
# plot(life_expectancy ~ population, data = life_exp4)
plot(life_expectancy ~ log(population), data = life_exp4)
```

```{r}
plot(life_expectancy ~ thinness_5_9_years, data = life_exp4)
```

```{r}
plot(life_expectancy ~ schooling, data = life_exp4)
```

```{r}
# plot(life_expectancy ~ infant_deaths, data = life_exp4)
plot(life_expectancy ~ log(infant_deaths + 1), data = life_exp4)
```

```{r}
plot(life_expectancy ~ diphtheria, data = life_exp4)
```

```{r}
# plot(life_expectancy ~ gdp, data = life_exp4)
plot(life_expectancy ~ log(gdp), data = life_exp4)
```

```{r}
plot(life_expectancy ~ thinness_1_19_years, data = life_exp4)
```

```{r}
plot(life_expectancy ~ income_composition_of_resources, data = life_exp4)
```

### Standardized residuals vs predictors - confirming log transformations are appropriate

```{r}
# constant variance satisfied - no transformation needed
plot(life_exp4$adult_mortality, rstandard(mlr_model),xlab = "Adult Mortality", ylab = "Standardized Residuals")

```

```{r}
plot (life_exp4$infant_deaths, rstandard(mlr_model),xlab = "infant death", ylab = "Standardized Residuals")

# constant variance satisfied - transformation needed
plot (log(life_exp4$infant_deaths), rstandard(mlr_model),xlab = "infant deaths", ylab = "Standardized Residuals")
```

```{r}
# constant variance satisfied - no transformation needed
plot(life_exp4$hepatitis_b, rstandard(mlr_model),xlab = "hepatitis_b", ylab = "Standardized Residuals")

```

```{r}
plot(life_exp4$measles, rstandard(mlr_model),xlab = "measles", ylab = "Standardized Residuals")

# constant variance satisfied - transformation needed
plot(log(life_exp4$measles), rstandard(mlr_model),xlab = "Measles", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied - no transformation needed
plot(life_exp4$bmi, rstandard(mlr_model),xlab = "bmi", ylab = "Standardized Residuals")

```

```{r}

plot(life_exp4$under_five_deaths, rstandard(mlr_model),xlab = "under_five_deaths", ylab = "Standardized Residuals")

# constant variance satisfied - transformation needed
plot(log(life_exp4$under_five_deaths), rstandard(mlr_model),xlab = "under five deaths", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied - no transformation needed
plot(life_exp4$polio, rstandard(mlr_model),xlab = "polio", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied - no transformation needed
plot(life_exp4$diphtheria, rstandard(mlr_model),xlab = "diphtheria", ylab = "Standardized Residuals")

```

```{r}
plot(life_exp4$hiv_aids, rstandard(mlr_model),xlab = "hiv_aids", ylab = "Standardized Residuals")

# constant variance satisfied - log transformation needed
plot(log(life_exp4$hiv_aids), rstandard(mlr_model),xlab = "hiv_aids", ylab = "Standardized Residuals")

```

```{r}
plot(life_exp4$gdp, rstandard(mlr_model),xlab = "gdp", ylab = "Standardized Residuals")

# constant variance satisfied - log transformation needed
plot(log(life_exp4$gdp), rstandard(mlr_model),xlab = "gdp", ylab = "Standardized Residuals")

```

```{r}

plot(life_exp4$population, rstandard(mlr_model),xlab = "Population", ylab = "Standardized Residuals")

# constant variance satisfied - transformation needed
plot(log(life_exp4$population), rstandard(mlr_model),xlab = "Population", ylab = "Standardized Residuals")
```

```{r}
# constant variance satisfied - no transformation needed
# mislabeled at thinness_1_19_years in dataset, actually represents 10-19 years
plot(life_exp4$thinness_1_19_years, rstandard(mlr_model),xlab = "thinness_10_19", ylab = "Standardized Residuals")

```

```{r}
# constant variance satisfied - no transformation needed
plot(life_exp4$thinness_5_9_years, rstandard(mlr_model),xlab = "thinness_5_9_years", ylab = "Standardized Residuals")
```

```{r}
# constant variance satisfied - no transformation needed
plot(life_exp4$income_composition_of_resources, rstandard(mlr_model),xlab = "Income Composition of Resources", ylab = "Standardized Residuals")
```

```{r}
# constant variance satisfied - no transformation needed
plot(life_exp4$schooling, rstandard(mlr_model),xlab = "Schooling", ylab = "Standardized Residuals")
```

### Interaction plots - interactions between status and all predictors

```{r}
ggplot(life_exp4, aes(adult_mortality, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(log(infant_deaths + 1), life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(hepatitis_b, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(log(measles+ 1), life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(bmi, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(bmi, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(log(under_five_deaths +1), life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(polio, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(diphtheria, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(log(hiv_aids), life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(gdp, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(population, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(thinness_1_19_years, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(thinness_5_9_years, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(income_composition_of_resources, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

```{r}
ggplot(life_exp4, aes(schooling, life_expectancy, colour = status)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE)
```

### Initial full model with transformations and interactions

```{r}
# Initial Full Model
lm_full <- lm(life_expectancy ~ status + adult_mortality + hepatitis_b + log(measles + 1) + bmi + log(under_five_deaths + 1) + polio + log(hiv_aids) + log(population) + thinness_5_9_years + schooling + log(infant_deaths + 1) + diphtheria + log(gdp) + thinness_1_19_years + income_composition_of_resources + adult_mortality:status + hepatitis_b:status + log(measles + 1):status + bmi:status + log(under_five_deaths + 1):status + polio:status + log(hiv_aids):status + log(population):status + thinness_5_9_years:status + schooling:status + log(infant_deaths + 1):status + diphtheria:status + log(gdp):status + thinness_1_19_years:status + income_composition_of_resources:status, data = life_exp4)

summary(lm_full)
```

### Trouble shooting initial full model - not full rank

```{r}
# Checking if model is full rank
X <- model.matrix(lm_full)
qrX <- qr(X)
qrX$rank
ncol(X)

# 31 < 32 the rank is smaller than the number of columns so our model is not full rank. 
```

```{r}
# use alias to identify terms that are perfectly colinear - interaction between status and log(hiv_aids)
alias(lm_full)
```

```{r}
# alias coefficients table
summary(lm_full)$aliased

```

### Create new model with redundant interaction term removed

```{r}
lm_full2 <- lm(life_expectancy ~ status + adult_mortality + hepatitis_b + log(measles + 1) + bmi + log(under_five_deaths + 1) + polio + log(hiv_aids) + log(population) + thinness_5_9_years + schooling + log(infant_deaths + 1) + diphtheria + log(gdp) + thinness_1_19_years + income_composition_of_resources + adult_mortality:status + hepatitis_b:status + log(measles + 1):status + bmi:status + log(under_five_deaths + 1):status + polio:status + log(population):status + thinness_5_9_years:status + schooling:status + log(infant_deaths + 1):status + diphtheria:status + log(gdp):status + thinness_1_19_years:status + income_composition_of_resources:status, data = life_exp4)
summary(lm_full2)
```

### Checking if Box Cox Transformation is useful

```{r}
# Box-Cox Transformation not needed because confidence interval contains 1
summary(powerTransform(lm_full))
boxcox(lm_full2, lambda=seq(0.7, 3, by=0.05))
```

### Model Selection Methods

```{r}
# Akaike Information Criterion (AIC) - step function
model_step <- step(lm_full2, trace = 0)
summary(model_step)
```

```{r}
# model with only 4 interactions run in dredge
lm_full3 <- lm(life_expectancy ~ status + adult_mortality + hepatitis_b + log(measles + 1) + bmi + log(under_five_deaths + 1) + polio + log(hiv_aids) + log(population) + thinness_5_9_years + schooling + log(infant_deaths + 1) + diphtheria + log(gdp) + thinness_1_19_years + income_composition_of_resources + adult_mortality:status + log(gdp):status + income_composition_of_resources:status + schooling:status, data = life_exp4)
```

```{r}
# AIC - Dredge function
options(na.action = "na.fail")
dredge_mod <- dredge(lm_full3, rank = "AIC",
                       subset = 
                         dc("adult_mortality", "adult_mortality:status") &
                         dc("status", "adult_mortality:status") &
                         dc("log(gdp)", "log(gdp):status") &
                         dc("status", "log(gdp):status") &
                         dc("income_composition_of_resources", "income_composition_of_resources:status") &
                         dc("status", "income_composition_of_resources:status") &
                         dc("schooling", "schooling:status") &
                         dc("status", "schooling:status"))
dredge_mod
```

```{r}
# extracting best model from dredge
dredge_mod_best <- get.models(dredge_mod, subset = 1)[[1]]
summary(dredge_mod_best)
```

### Final Full Model - same from step and dredge

```{r}
# Final Multiple Regression Model
lm_final <- lm(life_expectancy ~ status + adult_mortality + hepatitis_b + log(measles + 1) + log(hiv_aids) + status:adult_mortality + status:income_composition_of_resources, data = life_exp4)
summary (lm_final)
```

### Model Assumptions

```{r}
# standardized residuals
r <- rstandard(lm_final)

# fitted values
yhat <- fitted(lm_final)

# residual plot - constant variance satisfied
plot(yhat, r,
     xlab = "Fitted values",
     ylab = "Standardized residuals",
     main = "Standardized Residuals vs Fitted")
abline(h = 0, lty = 2)

# QQ plot - normality possibly satisfied
qqnorm(r)
qqline(r)

# Shapiro-Wilk - confirms normality satisfied
shapiro.test(resid(lm_final))

```

### Unusual Values

```{r}
# identifying outliers
ind <- which(abs(r) > 2)
life_exp4[ind, ]

# leverage
p <- 19
n <- nrow(life_exp4)
plot(hatvalues(lm_final), rstandard(lm_final),
     xlab = "Leverage",
     ylab = "Standardized Residuals")
abline(v = 2*(p+1)/n, lty=2)
abline(h = c(-2,2), lty = 2)

# identifying high leverage points
ind_lev <- which(hatvalues(lm_final) > 2*(p+1)/n)
life_exp4[ind_lev,]
```

### Interaction plots for status and adult mortality - investigating group of countries

```{r}
# all countries included
ggplot(life_exp4, aes(adult_mortality, life_expectancy, colour = status)) +
  geom_point(size = 0.7, alpha = 0.6) +
  geom_smooth(method = "lm") +
  labs(x = "Adult mortality",
       y = "Life expectancy",
       title = "Interaction plot with all countries")

# identifying group with low adult mortality, low life expectancy
ind <- which(life_exp4$adult_mortality < 45) 
life_exp4[ind,]

# group removed
life_exp5 <- life_exp4[-ind,] 

# plot with 14 countries removed
ggplot(life_exp5, aes(adult_mortality, life_expectancy, colour = status)) +
  geom_point(size = 0.7, alpha = 0.6) +
  geom_smooth(method = "lm") +
  labs(x = "Adult mortality",
       y = "Life expectancy",
       title = "Interaction plot with 15 countries removed")
```

### Looking into inconsistencies within data set

```{r}
# measles defined as number of cases per 1000 people
# 17 countries with counts > 1000
life_exp4 |>
  filter(measles > 1000)

max(life_exp4$measles)

# HIV/AIDS defined as deaths per 1000 live births, but in dataset as decimals
max(life_exp4$hiv_aids)
min(life_exp4$hiv_aids) # 0.1 deaths per 1000 births? Or 0.1 = 100/1000 deaths?

life_exp4 |>
  filter(status == "Developed")

life_exp4 |>
  filter(hiv_aids < 0.2) # 74 obs have hiv_aids = 0.1
```

```{r}
# model summary table
model_summary <- data.frame(
  Beta_0_8 = c(
    "Intercept",
    "Status: Developing",
    "Adult mortality",
    "Hepatitis B",
    "log(Measles + 1)",
    "log(HIV/AIDS)",
    "Income composition of resources",
    "Status: Developing × Adult mortality",
    "Status: Developing × Income composition of resources"
  ),
  Estimate = c(
    79.623404,
    -30.305370,
    -0.084900,
    0.027289,
    -0.146168,
    -1.314800,
    2.878530,
    0.068033,
    27.424461
  ),
  Std_error = c(
    17.678270,
    17.817764,
    0.021659,
    0.009715,
    0.074546,
    0.275648,
    18.822520,
    0.021897,
    19.002093
  ),
  t_value = c(
    4.504,
    -1.701,
    -3.920,
    2.809,
    -1.961,
    -4.770,
    0.153,
    3.107,
    1.443
  ),
  p_value = c(
    "1.55e-05",
    "0.091540",
    "0.000147",
    "0.005795",
    "0.052204",
    "5.20e-06",
    "0.878708",
    "0.002356",
    "0.151539"
  ),
  Significance = c(
    "*",
    ".",
    "*",
    "**",
    ".",
    "*",
    "",
    "**",
    ""
  )
)

model_summary
```

```{r}

```
